---
title: "TEMPLATE - alternative chart types practice"
author: "Zach Loo"
date: 2026-03-04
toc: true
---

::: {.callout-note}
This template accompanies the [alternative chart types practice slides](https://eds-240-data-viz.github.io/course-materials/lecture-slides/SLIDES-alt-charts.html#/title-slide){target="_blank}. Please be sure to cross-reference the slides, which contain important information and additional context!
:::

# Waffle Chart (Sasquatch sightings)

- Typefaces used: Ultra & Josefin Sans
- Colors used: "#357266", "#FFB813", "#983A06", "#005F71", "#757473", "#EFEFEF"

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#| fig-align: "center"
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                    setup                                 ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#..........................load packages.........................
library(tidyverse)
library(waffle)
library(showtext)

#..........................import data...........................
bigfoot <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-09-13/bigfoot.csv')

glimpse(bigfoot)
font_add_google('Emilys Candy', 'candy')
```

```{r fig.width=6, fig.height=3}
showtext_auto()
# Count by season, handle NAs
season_counts <- bigfoot |>
  filter(!is.na(season), season != 'Unknown') |>
  count(season) |>
  tibble::deframe()  # named vector required by waffle

# Season colors
season_colors <- c(
  "Spring" = "#78C679",
  "Summer" = "#F4D03F", 
  "Fall"   = "#E67E22",
  "Winter" = "#85C1E9"
)

bigfoot |>
  filter(!is.na(season), season != "Unknown") |>
  count(season) |>
  ggplot(aes(fill = season, values = n)) +
  geom_waffle(n_rows = 50, size = 0.5, colour = "white", flip = FALSE) +
  scale_fill_manual(values = c(
    "Spring" = "#78C679",
    "Summer" = "#F4D03F",
    "Fall"   = "#E67E22",
    "Winter" = "#85C1E9")) +
    labs(title = 'I Love Bigfeet') +
    coord_equal() +
    theme_void() +
    theme(
        legend.position = 'bottom',
        legend.title = element_blank(),
        plot.title = element_text(family = 'candy', hjust = 0.5, size = 30)
    )
```

```{r}
ggsave(here::here('week9', 'bigfoot.png'), width = 6, height = 3, dpi = 150)
```

# Bump chart (highest paying occupations)

- Typefaces used: Passion One & Oxygen
- Colors used: `viridisLite::magma()` & "#2A114E", "#6D6B71", "#FFF8F4"

```{r}
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                    setup                                 ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#..........................load packages.........................
library(tidyverse)
library(ggbump)
library(ggtext)
library(showtext)

#..........................import data...........................
jobs <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-03-05/jobs_gender.csv")
```

# Stream graph (energy production)

- Typefaces used: Exo 2 & Open Sans
- Colors used: `paletteer::paletteer_d("MetBrewer::Archambault", n = 7)`

```{r}
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                    setup                                 ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#..........................load packages.........................
library(tidyverse)
library(ggstream)
library(ggtext)
library(showtext)

#..........................import data...........................
capacity <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2022/2022-05-03/capacity.csv")
```











