---
title: "TEMPLATE - visualizing numerical relationships"
author: "Your Name"
date: 2026-01-28
toc: true
editor_options: 
  chunk_output_type: inline
---

::: {.callout-note}
This template follows [relationships slides](https://eds-240-data-viz.github.io/course-materials/lecture-slides/SLIDES-relationships.html#/title-slide){target="_blank"}. Please be sure to cross-reference the slides, which contain important information and additional context!
:::

## Setup

Data are downloaded from [DataOne](https://search.dataone.org/view/https%3A%2F%2Fpasta.lternet.edu%2Fpackage%2Fmetadata%2Feml%2Fknb-lter-hbr%2F208%2F14){target="_blank} (find instructions for getting the download link on the [`{metajam}` package README](https://github.com/NCEAS/metajam?tab=readme-ov-file#how-to-get-the-url-to-your-dataset-of-interest-){target="_blank}).

```{r}
#| eval: false
#| echo: true
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                    setup                                 ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#..........................load packages.........................
library(metajam) 
library(tidyverse)
library(ggExtra)
library(ggdensity)

#...................download data from DataOne...................
# you only need to do this once (then I recommend commenting it out)!
# metajam::download_d1_data("https://cn.dataone.org/cn/v2/resolve/https%3A%2F%2Fpasta.lternet.edu%2Fpackage%2Fdata%2Feml%2Fknb-lter-hbr%2F208%2F14%2F024b6acc5cb2e03a14fff5558bbffc0c",
#                  path = here::here("week4")) 

#  ~ NOTE: You should rename the downloaded folder to 'data/' so that it's automatically ignored by your repo's .gitignore! ~

#....................read in downloaded files....................
#stream_chem_all <- metajam::read_d1_files(here::here("week4", "data"))

#........................get the data file.......................
# stream_chem_data <- stream_chem_all$`HubbardBrook_weekly_stream_chemistry_1963-2025.csv`

stream_chem_all <- readr::read_csv(here::here("week4", "data", "HubbardBrook_weekly_stream_chemistry_1963-2025.csv"))
```

## Scatter plots

### Basic scatter plot

```{r}
#| eval: true
#| echo: true
#| warning: false
#.......................basic scatter plot.......................
p1 <- stream_chem_all |> 
  filter(waterYr == 2024) |> 
  ggplot(aes(DOC, pH)) +
  geom_point(alpha = 0.4)
```

### Alt 1: add a rug plot

```{r}
#| eval: true
#| echo: true
#| warning: false
#.......................scatter + rug plot.......................
p1 +
  geom_rug()
```

### Alt 2: marginal plot alternatives using `{ggExtra}`

```{r}
#| eval: true
#| echo: true
#| warning: false
#................add marginal plots with {ggExtra}...............
# also try "density", "boxplot"
ggMarginal(p1, type = "histogram")
```

### Alt 3: scatter + marginal plot with 2+ groups

```{r}
#| eval: true
#| echo: true
#| warning: false
#............scatter plot with points colored by site............
p2 <- stream_chem_all |> 
  filter(waterYr == 2024) |> 
  ggplot(aes(x = DOC, y = pH, color = site)) + 
  geom_point(alpha = 0.5) +
  theme(legend.position = 'bottom')

p2

#..................with marginal density plot....................
ggMarginal(p2, type = "density", groupFill = TRUE, groupColour = TRUE)
```

## Trend lines

### Default (<1000 data points): LOESS method

```{r}
#| eval: true
#| echo: true
#......................add a smoothed line.......................
stream_chem_all |> 
  filter(waterYr %in% c(2014:2024)) |> 
  filter(site == "W5") |> 
  ggplot(aes(DOC, pH)) +
  geom_point() +
  geom_smooth()
```

### Line of best fit

```{r}
#| eval: true
#| echo: true
#........................add a trend line........................
stream_chem_all |> 
  filter(waterYr %in% c(2014:2024)) |>  
  filter(site == "W5") |> 
  ggplot(aes(x = DOC, y = pH)) + 
  geom_point(alpha = 0.5) +
  geom_smooth(method = 'lm', se = FALSE)
```

## Visualizing a third numeric variable

### Bubble charts

- be mindful: it's challenging to interpret 3 numeric variables
    
```{r}
#| eval: true
#| echo: true
#| warning: false
#..........................bubble chart..........................
stream_chem_all |> 
  filter(waterYr == 2024) |> 
  ggplot(aes(x = DOC, y = pH, color = site, size = Al_ICP)) + 
  geom_point(alpha = 0.5) +
  labs(x = "DOC (mg/L)", size = "Al (mg/L)", color = "Site")
```

- adjust size range of bubbles
- IMPORTANT: don't scale size by radius (can be deceptive)

```{r}
#| eval: true
#| echo: true
#| warning: false
#| layout-ncol: 2
#.......................scale size by area.......................
stream_chem_all |> 
  filter(waterYr == 2024) |> 
  ggplot(aes(x = DOC, y = pH, color = site, size = Al_ICP)) + 
  geom_point(alpha = 0.5) +
  scale_size(range = c(1, 10)) +
  labs(x = "DOC (mg/L)", size = "Al (mg/L)", color = "Site")

#...................DON'T scale size by radius...................
# stream_chem_data |> 
#   filter(waterYr == 2024) |> 
#   ggplot(aes(x = DOC, y = pH, color = site, size = Al_ICP)) + 
#   geom_point(alpha = 0.5) +
#   scale_radius(range = c(1, 10)) +
#   labs(x = "DOC (mg/L)", size = "Al (mg/L)", color = "Site")
```

### Use color for third numeric variable

```{r}
#| eval: true
#| echo: true
#| warning: false
#...................color for a third variable...................
stream_chem_all |> 
  filter(waterYr == 2024) |> 
  ggplot(aes(x = DOC, y = pH, color = Al_ICP)) + 
  geom_point(alpha = 0.5, size = 2) +
  scale_color_viridis_c() +
  labs(x = "DOC (mg/L)")
```

### But oftentimes best to just create two different plots

```{r}
#| eval: true
#| echo: true
#| warning: false
#| layout-ncol: 2
#......................effect of DOC on pH.......................
stream_chem_all |> 
  filter(waterYr == 2024) |> 
  ggplot(aes(x = DOC, y = pH, color = site)) + 
  geom_point(alpha = 0.5)

#....................effect of DOC on Al_ICP.....................
stream_chem_all |> 
  filter(waterYr == 2024) |> 
  ggplot(aes(x = Al_ICP, y = pH, color = site)) + 
  geom_point(alpha = 0.5)
```

## Overplotting

- if you have too many points, scatter plots can be ineffective

```{r}
#| eval: true
#| echo: false
#| warning: false
#.....................wayyyy too many points.....................
ggplot(stream_chem_all, aes(SO4, pH)) +
  geom_point()
```

### Initial strategies

```{r}
#| eval: true
#| echo: false
#| warning: false
#| layout-ncol: 3
#................smaller, more transparent points................
ggplot(stream_chem_all, aes(x = SO4, y = pH)) + 
  geom_point(alpha = 0.4, size = 0.5) 

#..........................add rug plot..........................
ggplot(stream_chem_all, aes(x = SO4, y = pH)) + 
  geom_point(size = 0.5, alpha = 0.3) +
  geom_rug()

#.........................color by group.........................
ggplot(stream_chem_all, aes(x = SO4, y = pH, color = site)) + 
  geom_point(size = 0.5, alpha = 0.3)
```

### Alt 1: 2d heatmaps

- e.g. like you're looking down on a histogram

```{r}
#| eval: true
#| echo: false
#| warning: false
#| layout-ncol: 2
#....................heatmap of 2d bin counts....................
ggplot(stream_chem_all, aes(x = SO4, y = pH)) +
  geom_bin2d() +
  scale_fill_viridis_c()

#..................heatmap with hexagonal bins...................
ggplot(stream_chem_all, aes(x = SO4, y = pH)) +
  geom_hex() +
  scale_fill_viridis_c() +
  guides(fill = guide_colorbar(title = "Count", 
                                barwidth = 1, barheight = 15))
```

### Alt 2: 2d density / contour plots

- e.g. like you're looking down on a density plot
- legend provides an estimate of the proportion of data points fall within a colored region (density of distribution of points sums to 1)
    - interpreting legend: 0-2% of of points fall within a 1x1 square in the darkest blue region, while 24-26% fall within a 1x1 square in the brightest yellow region

```{r}
#| eval: true
#| echo: false
#| warning: false
#..........................contour plot..........................
ggplot(stream_chem_all, aes(x = SO4, y = pH)) +
  geom_density_2d_filled()
```

### Alt 3: 2d density plots using `{ggdensity}`

- can be easier to interpret
- compute and plot the resulting highest density regions (HDRs)
- HDRs are computed to be the smallest such regions that bound that level of probability

```{r}
#| eval: true
#| echo: false
#| warning: false
#................2d density plot with {ggdensity}................n. 
ggplot(stream_chem_all, aes(x = SO4, y = pH)) +
  geom_hdr()
```
