---
title: "TEMPLATE - visualizing evolution"
author: "Your Name"
date: 2026-01-21
toc: true
---

::: {.callout-note}
This template follows the [visualizing evolution slides](https://eds-240-data-viz.github.io/course-materials/lecture-slides/SLIDES-evolution.html#/title-slide){target="_blank}. Please be sure to cross-reference the slides, which contain important information and additional context!
:::

## Setup

::: {.callout-note}
`{tidycensus}` functions (e.g. `get_estimates()`) internally call `Sys.getenv("CENSUS_API_KEY)` to retrieve your key from `.Renviron`. This means you don't need to explicitly call `Sys.getenv("CENSUS_API_KEY)` in your code -- `{tidycensus}` handles that for you!
:::

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                    setup                                 ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#.........................load libraries.........................
library(tidycensus)
library(tidyverse)
library(gghighlight)
library(scales)

#..........................import data...........................
lyme <- read_csv(here::here("week3", "data", "2023_CaseIncid-Lyme-Disease-Cases-by-State-or-Locality.csv"))

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                          wrangle lyme disease data                       ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#....................wrangle lyme disease data...................
lyme_clean <- lyme |> 
  
  rename(state = State) |> 
  
  # remove special character from State names that have it ----
  mutate(state = str_replace(string = state, pattern = "\x86", replacement = "")) |> 
  
  # remove total cases row ----
  filter(state != "U.S. Total") |> 
  
  # wide to long (tidy) years ----
  pivot_longer(cols = 2:17, names_to = "year", values_to = "reported_cases") |> 
  
  # reorder columns ----
  relocate(year, state, reported_cases) |> 
  
  # filter for years 2010 - 2020 ----
  filter(year %in% c(2010:2020))
  
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                          get & wrangle census data                       ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#........get pop estimates by state using tidycensus API.........
us_state_pop <- get_estimates(geography = "state", 
                              product = "population",
                              state = NULL, 
                              year = 2019) |> # request 2019 vintage for the 2010s (see documentation)
  filter(variable == "POP") |> 
  select(state = NAME, population = value) |> 
  filter(state != "Puerto Rico")

#........................write data to csv.......................
# OPTIONAL, but recommended in case you want to work offline, the API is down, etc. (you can then read in your saved data file rather than run the above code)
# write_csv(us_state_pop, file = here::here("week3", "data", "us_state_pop.csv"))  

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                            join lyme & pop dfs                           ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

lyme_pop <- left_join(lyme_clean, us_state_pop) |>  
  
  # add col with num of 100k people per state ----
  mutate(pop100k = population/100000) |> 
  
  # calculate num cases per 100k people (common way of reporting disease incidence) ----
  mutate(cases_per100k = reported_cases/pop100k) 
```

## Line graphs

- show evolution of one or more numeric variables; similar to scatter plots but points are ordered (typically by x-axis value) and joined with straight line segments

```{r}
#| eval: true
#| echo: true
 #...........................line chart...........................
```

### Avoid spaghetti plots 

- highlight groups or values of interest to focus attention

```{r}
#| eval: true
#| echo: true
#| warning: false
#| message: false

#.......................highlight on group.......................
ggplot(lyme_pop, aes(x = year, y = cases_per100k, group = state)) +
  geom_line() 

#..................highlight based on criteria...................
ggplot(lyme_pop, aes(x = year, y = cases_per100k, group = state)) +
  geom_line() 
```

### Note 1: ggplot cuts the y-axis if 0 isn’t relevant

- line graphs encode data **by position** and **not length** -- cutting the y-axis is generally considered acceptable

```{r}
#| eval: true
#| echo: true
#| layout-ncol: 2

#..................ggplot cuts y-axis by default.................
lyme_pop |> 
  filter(year %in% c(2013:2019)) |> 
  filter(state == "Vermont") |> 
  ggplot(aes(x = year, y = cases_per100k, group = state)) +
  geom_line()

#....................force y-axis origin at 0....................
lyme_pop |> 
  filter(year %in% c(2013:2019)) |> 
  filter(state == "Vermont") |> 
  ggplot(aes(x = year, y = cases_per100k, group = state)) +
  geom_line() +
  scale_y_continuous(limits = c(0, NA))
```

### Note 2: Consider aspect ratio (height:width ratio)

::: {.callout-tip}
Temporarily adding tick marks (breaks) can help to visualize changes to your plot's aspect ratio. Always remove breaks when finalizing your plot. 
:::

```{r}
#| eval: true
#| echo: true
#..........default aspect ratio (with added tick marks)..........
lyme_pop |> 
  filter(state == "Vermont") |> 
  ggplot(aes(x = year, y = cases_per100k, group = state)) +
  geom_line() 
```

## Area charts

- similar to line chart; indicates change in a numeric variable through time, but the space between x-axis and line is shaded in

```{r}
#| eval: true
#| echo: true
#...........area chart of lyme disease incidence in NJ...........
lyme_pop |> 
  filter(state == "New Jersey") 
```

### Alternative: stacked area charts 

- show evolution of the whole + relative proportions of each group

```{r}
#| eval: true
#| echo: true
#.......................stacked area chart.......................
lyme_pop |> 
  filter(state %in% c("Maine", "Rhode Island", "New Hampshire", "Vermont")) 
```

### Alternative: proportional stacked area chart

- plot percentage contributions rather than absolute numbers on y-axis

```{r}
#| eval: true
#| echo: true
#................proportional stacked area chart.................
lyme_pop |> 
  filter(state %in% c("Maine", "Rhode Island", "New Hampshire", "Vermont")) |> 
  ggplot(aes(x = year, y = cases_per100k, group = state, fill = state)) +
  geom_area() 
```

### Note 1: stacked area charts aren't great for looking at evolution of individual groups

- the stacked area chart and faceted line graph both show the same data

```{r}
#| eval: true
#| echo: true
#| layout-ncol: 2
#.......................stacked area chart.......................
lyme_pop |> 
  filter(state %in% c("Maine", "Rhode Island", "New Hampshire", "Vermont")) |> 
  ggplot(aes(x = year, y = cases_per100k, group = state, fill = state)) +
  geom_area()

#........same data, but as a line chart & faceted by state.......
lyme_pop |> 
  filter(state %in% c("Maine", "Rhode Island", "New Hampshire", "Vermont")) |> 
  ggplot(aes(x = year, y = cases_per100k, group = state, color = state)) +
  geom_line() +
  facet_wrap(~state)
```

